Shapes analysis for time series.

### 2024-05-22 NEURIPS24/appendix.tex
Changes

\paragraph{TS-LDDMMM parameters.}
\begin{itemize}
  \item \textbf{Before exposure:} The velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,150,1,2)$. The varifold loss is the sum of three varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3): $\left((5,2,5,1),(2,1,2,0.6),(1,0.6,1,0.6)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.3.
  \item \textbf{Before/after exposure:} The velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,220,1,2)$. The varifold loss is the sum of four varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3, Varifold 4): $\left((30,2,30,1),(5,2,5,1),(2,1,2,0.6),(1,0.1,1,0.1)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.3.
\end{itemize}

\paragraph{LDDMMM parameters.} Note that varifold losses are unchanged between TS-LDDMM and LDDMM. Compared to TS-LDDMM, the convergence of LDDMM is more sensitive to the maximum stepsize $\eta_m$, which must remain small for LDDMM to guarantee the convergence.
\begin{itemize}
  \item \textbf{Before exposure:} The velocity field kernel $K_G$ is an anysotropic Gaussian kernel with parameters $\sigma_{T} =150$ for the time dimension and $\sigma_x = 2$ for space dimensions. 
  The varifold loss is the sum of three varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3): $\left((5,2,5,1),(2,1,2,0.6),(1,0.6,1,0.6)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.01.
  \item \textbf{Before/after exposure:} The velocity field kernel $K_G$ is an anysotropic Gaussian kernel with parameters $\sigma_{T} =220$ for the time dimension and $\sigma_x = 2$ for space dimensions.
  The varifold loss is the sum of four varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3, Varifold 4): $\left((30,2,30,1),(5,2,5,1),(2,1,2,0.6),(1,0.1,1,0.1)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.01.
\end{itemize}

@@

\paragraph{TS-LDDMM and LDDMM}

\subsection{Protocol}

\subsection{Results}

\begin{table}[hbt!]
  \centering
  \begin{tabular}{lcccc}
    \toprule
    \multirow[c]{2}{*}{\textbf{Methods}} & \multicolumn{4}{c}{\textbf{Test F1-score}} \\
    \cline{2-5}
     &  \textbf{Regular} & \textbf{30 \% dropped} &  \textbf{50 \% dropped} & \textbf{70 \% dropped} \\
    \midrule
    RNN (1999) & $0.64 \pm 0.21$ & $0.53 \pm 0.23$ & $0.48 \pm 0.21$ & $0.44 \pm 0.21$ \\
    LSTM (1997) & $0.61 \pm 0.29$ & $0.57 \pm 0.29$ & $0.53 \pm 0.25$ & $0.51 \pm 0.29$ \\
    GRU (2014) & $0.71 \pm 0.26$ & $0.68 \pm 0.28$ & $0.66 \pm 0.28$ & $\underline{0.59 \pm 0.28}$ \\
    MTAN (2021) & $0.59 \pm 0.28$ & $0.58 \pm 0.28$ & $0.54 \pm 0.29$ & $0.51 \pm 0.28$ \\
    MIAM (2022) & $0.48 \pm 0.35$ & $0.42 \pm 0.33$ & $0.47 \pm 0.31$ & $0.35 \pm 0.31$ \\
    ODE-LSTM (2020) & $0.63 \pm 0.24$ & $0.57 \pm 0.25$ & $0.51 \pm 0.24$ & $0.45 \pm 0.23$ \\
    Neural SDE (2019) & $0.48 \pm 0.28$ & $0.47 \pm 0.26$ & $0.45 \pm 0.27$ & $0.45 \pm 0.25$ \\
    Neural LNSDE (2024) & $0.7 \pm 0.27$ & $0.68 \pm 0.29$ & $\underline{0.67 \pm 0.25}$ & $\mathbf{0.66 \pm 0.23}$ \\
    LDDMM (2008) & $\underline{0.72 \pm 0.2}$ & $\underline{0.7 \pm 0.21}$ & $0.57 \pm 0.25$ & $0.4 \pm 0.25$ \\
    %\cline{1-5}
    TS-LDDMM (ours) & $\mathbf{0.83 \pm 0.18}$ & $\mathbf{0.8 \pm 0.18}$ & $\mathbf{0.7 \pm 0.26}$ & $0.51 \pm 0.27$ \\
    \bottomrule
  \end{tabular}  
\end{table}

\subsection{Appendix for experiment: Classification benchmark on regularly sampled datasets}
\subsection{Settings}
\label{appendix: settings_classification}

In this section, we compare classification performances of TS-LDDMM with other state-of-the-art methods coming from shape analysis on 15 shape-based datasets of time-series.

>>>>

\paragraph{TS-LDDMM parameters.}
\begin{itemize}
  \item \textbf{Before exposure:} The velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,150,1,2)$. The varifold loss is the sum of three varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3): $\left((5,2,5,1),(2,1,2,0.6),(1,0.6,1,0.6)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.3.
  \item \textbf{Before/after exposure:} The velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,220,1,2)$. The varifold loss is the sum of four varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3, Varifold 4): $\left((30,2,30,1),(5,2,5,1),(2,1,2,0.6),(1,0.1,1,0.1)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.3.
\end{itemize}

\paragraph{LDDMM parameters.} Note that varifold losses are unchanged between TS-LDDMM and LDDMM. Compared to TS-LDDMM, the convergence of LDDMM is more sensitive to the maximum stepsize $\eta_m$, which must remain small for LDDMM to guarantee the convergence.
\begin{itemize}
  \item \textbf{Before exposure:} The velocity field kernel $K_G$ is an anysotropic Gaussian kernel with parameters $\sigma_{T} =150$ for the time dimension and $\sigma_x = 2$ for space dimensions. 
  The varifold loss is the sum of three varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3): $\left((5,2,5,1),(2,1,2,0.6),(1,0.6,1,0.6)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.01.
  \item \textbf{Before/after exposure:} The velocity field kernel $K_G$ is an anysotropic Gaussian kernel with parameters $\sigma_{T} =220$ for the time dimension and $\sigma_x = 2$ for space dimensions.
  The varifold loss is the sum of four varifolds to capture shapes variations at different scales with parameters: (Varifold 1,Varifold 2,Varifold 3, Varifold 4): $\left((30,2,30,1),(5,2,5,1),(2,1,2,0.6),(1,0.1,1,0.1)\right)$ and the mapper $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x})$. The optimizer has 800 steps with a maximum stepsize $\eta_M$ of 0.01.
\end{itemize}

@@

For each method and dataset, the learning rate, the hidden vector dimensions, and the number of layers are optimized to minimize the CrossEntropy loss on a validation set using the \texttt{Ray}~\footnote{\url{https://github.com/ray- project/ray}} Python library. 
The learning rate varies from $10^{-4}$ to $10^{-1}$ using log uniform search, the hidden vector dimension ranges from ${16, 32, 64, 128}$ using grid search, and the number of layers ranges from ${1, 2, 3, 4}$ using grid search. The batch size was selected from ${16, 32, 64, 128}$ according to the size of the dataset. All methods were trained for 100 epochs, and the best method was selected based on the lowest validation loss.

\paragraph{TS-LDDMM and LDDMM}
Representations learned with TS-LDDMM or LDDMM are fed to a Support Vector Classifier (\href{https://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html#sklearn.svm.SVC}{SVC}) from \texttt{scikit-learn}~\footnote{\url{https://scikit-learn.org/stable/}}. All SVC's hyperparameters are set to default except the regularization term C, which is set through grid search on a validation set with the macro f1-score~\footnote{\url{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.f1_score.html}}.

To learn TS-LDDMM (resp. LDDMM) representations, the velocity field kernel $K_G$ is set to $ (c_0,c_1,\sigma_{T,0},\sigma_{T,1},\sigma_x) = (1,0.1,0.33\bar{l},1,n_d)$, (resp. $ (\sigma_{T},\sigma_x) = (0.33\bar{l},n_d)$) where $\bar{l}$ is the average time series length and $n_d$ the number of dimensions. For both methods and all datasets, the varifold loss kernels $(k_{pos},k_{dir})$ are identical and set to $(\sigma_{\pos,t}, \sigma_{\pos,t}, \sigma_{\dir,t}, \sigma_{\dir,x}) = (2,n_d,2,n_d)$. For TS-LDDMM (resp. LDDMM), the optimizer is set with 400 epochs (resp. 400) and a maximum learning rate $\eta_M = 0.1$ (resp. $\eta_M = 0.01$). In all cases, the initial reference graph is selected in the dataset as a time series with the median length.

\subsection{Protocol}
In this experiment, we investigate the robustness to missing samples and the classification performance of TS-LDDMM compared to Neural ODEs on 15 datasets described in \Cref{appendix:classification_dataset}. For fairness between methods of different architectures, the evaluation protocol on each dataset and method is as follows: 
\begin{enumerate}
  \item The dataset is spilt in train 75\%, validation 15\%, and test 15\%.
  \item hyperparameters are tuned with train and validation sets and a missing rate of 0\%.
  \item For each missing rate in [0\%,30\%,50\%,70\%]
    \begin{itemize}
      \item Remove samples in time series in the train and test sets according to the missing rate and the drop procedure described in~\cite{kidger2020neural}.
      \item Train the model on the train set
      \item Evaluate the macro f1-score on the test set
    \end{itemize}
\end{enumerate}

\subsection{Results}
In this experiment, we investigate the robstuness to missing samples and the classification performance of TS-LDDMM representations. We compare TS-LDDMM with LDDMM and 8 neural ODEs networks. Performances are evaluated in terms of average macro f1-score and rank on four different regimes of missing rate 0\%,30\%,50\%, and 70\%. Results are aggregated in \Cref{table:robstuness}.

On three out of four regimes (0\%,30\%, and 50\%) TS-LDDMM classifier is the best performer in terms of f1-score and rank. For missing rates of 0\% and 30\%, the score increases by 10\% compared to the second-best performer, LDDMM. However, LDDMM is not the second-best performer in rank (Neural LNSDE), showing its sensitivity to parameterization, unlike TS-LDDMM, which remains consistent. Performances of Neural LNSDE remain constant with the increase of the missing rate as observed in~\cite{oh2024stable}, and it becomes the best performer for missing rate 70\%. The decrease in TS-LDDMM performances with the increasing missing rate is due to the varifold loss, which poorly approximates the time series shape. Other losses might be more relevant for high missing rates.

Overall, TS-LDDMM is a relevant and consistent shape-based representation for irregularly sampled multivariate time series for missing rates up to 50\% .

\begin{table}[hbt!]
  \centering
  \caption{Comparison of average macro f1-score and rank as the sample dropping rate increases. \textbf{First} \& \underline{second} best performers. TS-LDDMM is the best performer on three out of four regimes.}
  \label{table:robstuness}
  \resizebox{\columnwidth}{!}{%
  \begin{tabular}{lcccccccc}
    \toprule
    \multirow[c]{2}{*}{\textbf{Methods}} & \multicolumn{2}{c}{\textbf{Regular}} & \multicolumn{2}{c}{\textbf{30 \% dropped}} &  \multicolumn{2}{c}{\textbf{50 \% dropped}} & \multicolumn{2}{c}{\textbf{70 \% dropped}} \\
    \cline{2-9}
     &  \textbf{F1-score} & \textbf{Rank} &  \textbf{F1-score} & \textbf{Rank} &  \textbf{F1-score} & \textbf{Rank} &  \textbf{F1-score} & \textbf{Rank} \\
    \midrule
    RNN (1999) & $0.64 \pm 0.21$  & 6.2 & $0.53 \pm 0.23$  & 6.6 & $0.48 \pm 0.21$  & 7.2 & $0.44 \pm 0.21$  & 6.07 \\
    LSTM (1997) & $0.61 \pm 0.29$  & 6.0 & $0.57 \pm 0.29$  & 6.27 & $0.53 \pm 0.25$  & 6.07 & $0.51 \pm 0.29$  & 5.27 \\
    GRU (2014) & $0.71 \pm 0.26$  & 4.2 & $0.68 \pm 0.28$  & 4.27 & $0.66 \pm 0.28$  & 3.73 & $\underline{0.59 \pm 0.28}$  & \underline{3.67} \\
    MTAN (2021) & $0.59 \pm 0.28$  & 7.13 & $0.58 \pm 0.28$  & 5.8 & $0.54 \pm 0.29$  & 5.33 & $0.51 \pm 0.28$  & 5.0 \\
    MIAM (2022) & $0.48 \pm 0.35$  & 6.93 & $0.42 \pm 0.33$  & 8.27 & $0.47 \pm 0.31$  & 6.93 & $0.35 \pm 0.31$  & 7.6 \\
    ODE-LSTM (2020) & $0.63 \pm 0.24$  & 6.0 & $0.57 \pm 0.25$  & 6.53 & $0.51 \pm 0.24$  & 7.27 & $0.45 \pm 0.23$  & 6.73 \\
    Neural SDE (2019) & $0.48 \pm 0.28$  & 7.67 & $0.47 \pm 0.26$  & 7.47 & $0.45 \pm 0.27$  & 7.13 & $0.45 \pm 0.25$  & 6.0 \\
    Neural LNSDE (2024) & $0.7 \pm 0.27$  & \underline{3.87} & $0.68 \pm 0.29$  & \underline{4.0} & $\underline{0.67 \pm 0.25}$  & \underline{3.53} & $\mathbf{0.66 \pm 0.23}$  & \textbf{2.47} \\
    LDDMM (2008) & $\underline{0.72 \pm 0.2}$  & 4.53 & $\underline{0.7 \pm 0.21}$  & 4.2 & $0.57 \pm 0.25$  & 5.0 & $0.4 \pm 0.25$  & 7.13 \\
    TS-LDDMM (ours) & $\mathbf{0.83 \pm 0.18}$  & \textbf{2.93} & $\mathbf{0.8 \pm 0.18}$  & \textbf{2.07} & $\mathbf{0.7 \pm 0.26}$  & \textbf{3.33} & $0.51 \pm 0.27$  & 5.67 \\
    \bottomrule
  \end{tabular}  
  %
  }
\end{table}

\subsection{Appendix for experiment: Classification benchmark on regularly sampled datasets}

In this section, we compare the classification performances of TS-LDDMM with other methods from shape analysis on 15 shape-based datasets of time series regularly sampled. TS-LDDMM outperforms other methods on 12 out of 15, highlighting its relevance for shape analysis when dealing with time series.

\subsection{Benchmark metohds}
\begin{itemize}
  \item SRV-based method: we include TCLR~\cite{heo2024logistic} a logistic regression on the tangent space of the Frechet mean with Square Root Velocity (SRV representation). We also include Shape-FPCA~\cite{wu2024shape} that encodes both the time series and its time parameterization. Both methods are implemented with the Python library \texttt{FDASRSF}~\footnote{https://fdasrsf-python.readthedocs.io/en/latest/}, which cannot handle multivariate time series.
  \item LDDMM-Based : TS-LDDMM (ours) and LDDMM~\cite{glaunes2008large}. Both methods learn representations by solving ODEs parametrized with Kernels. While both methods handle multivariate signals, TS-LDDMM is specifically designed for time series.
 \end{itemize}
\subsection{Settings}
\label{appendix: settings_classification}
